---
title: Converting LaTeX to Word using pandoc
author: ''
date: '2022-09-17'
slug: converting-latex-to-word-using-pandoc
categories:
  - R markdown
tags:
  - pandoc
description: ''
highlight: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, comment = NA, 
                      warning = FALSE, message = FALSE)
```

A few weeks ago I needed to convert an academic paper written using [LaTeX](https://en.wikibooks.org/wiki/LaTeX){target='_blank'} into a Microsoft Word document. My preferred way of doing this would be to use [pandoc](https://pandoc.org/){target='_blank'}. 

In the past I have used [pandoc](https://pandoc.org/){target='_blank'} in the command line. On checking my (reasonably) new laptop, it appeared that I didn't have access to it on the command line but I knew it had to be installed as part of the [RStudio](https://www.rstudio.com/){target='_blank'} install. 

This led me to check if there were any `pandoc()` functions within any of the packages and I found that, as I suspected, there are a bunch of them, mostly in the [rmarkdown package](https://rmarkdown.rstudio.com/docs/index.html){target='_blank'}. To demo, we can use it to check with `pandoc` is installed 

```{r}
library(rmarkdown)
pandoc_available()
```

and check which version is available.

```{r}
pandoc_version()
```

Now that we're happy we can access pandoc, we can use `pandoc_convert()` to do the actual conversion. My LaTeX document also used a [BibTeX](https://en.wikibooks.org/wiki/LaTeX/Bibliography_Management){target='_blank'} file and so the conversion could be a little more complicated. The command line syntax is explained nicely [here](https://waterprogramming.wordpress.com/2018/08/26/converting-latex-to-ms-word-docx-almost-perfectly/){target='_blank'} and we can implement it in `R` even more easily as we don't actually need to specify the BibTeX file (it is picked up from the LaTeX file). 

```{r, eval = FALSE}
pandoc_convert(input = "my_paper.tex", 
               output = "my_paper_converted.docx", 
               citeproc = TRUE)
```

In my case, the conversion worked really well and I will be trying this in the future to allow quick conversions between LaTeX (my preferred option for writing papers) to help with collaboration. 
